Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-07-26 manuscript.rmd
1. Revised text 2. Added table for sp code names 3. Filled up info on ELM-FATEs parameterisation table: soil vulnerability curves, soil hydro conductivity calculation and corresponding figures 4. added ksat values in parameter table

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
bibliography: references.bib
csl: new-phytologist.csl
---

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
```

@@

- Deep water access (WA~Deep~) is believed to be an important, drought-tolerance mechanism, but under-studied due to lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test its role in drought-tolerance strategies of a tropical forest community (Panama) with a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for 32 large, canopy tree species by linking species growth dynamics (xx trees, 6 censuses, 1990-2015) to VPD, $\Psi_{x,~z}$ by depth $z$in the whole-soil column, and a novel community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$) linking data on the same for 21 species with community-scale softer traits. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$). 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margin, but lower mean mortality rate. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for modeling the future of tropical species and forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora:2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic global vegetation models (DGVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith:2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012].

\
Trait-based plant ecology aims to predict impacts of environmental change on community composition and function using relevant plant traits as currency. To reach this goal it is imperative to establish relationships between plant's environment, traits and demographic outcomes, but this remains to be a challenge. Huge efforts have been devoted to linking same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in above-ground hydraulic traits such as proxies of hydraulic failure as suggested indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with high vulnerability to drought-induced embolism ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$), but counter examples also exist (Hoffmann 2011; Paddock 2013), indicating whether plants actually reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a rare, centennial drought (Texas, USA) found that deep rooted species died less despite being more vulnerable to xylem embolism (high $\Psi_{88,~Stem}$) and having narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2017]. A species sourcing deep water showed lower hydraulic failure and lower mortality in an extreme drought in an Italian Karst system even though it had narrower hydrulic safety margin and high $\Psi_{88,~Stem}$ than other two species limited to shallow-water[@Nardini:2015]. Chitra-Tarak et al. (2018) proposed a general model to link realised water availability to demographic outcomes through droughts at tree community-scale. This study found that species accessing deep-water in a tropical dry forest study system died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with expectation of deep water usage as an aquisitive strategy relying on stable and reliable water resource at depth, which exhausted in the multi-annual drought resulting in reduced survival, compared to persistent species associated with shallow-water that was relatively more available via infiltration during this drought. However, species estimates of water-access modelled by this study were not corroborated becuase of lack of data on rooting/water-sourcing depths, or related hydraulic traits.

\
Here, we obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column of a neo-tropical forest (BCI, Panama) with a locally parametarised and calibrated 1-D hydrological water balance model. We then estimated the effective rooting depths for xx large, canopy tree species as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015) with a novel inverse model. This model incorporated impact of atmospheric drought (VPD) on growth and new data on species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$), which we scaled-up to community level via scaling relationships with commonly available traits. We evaluated the inversely estimated ERD against exisiting data on plant and soil water isotopes ($\delta^2H$).

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a conservative-acquisitive trade-off axis. Specific hypotehses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper effective rooting depths (ERD) show acquisitive hydraulic architecture: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins; and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a tropical forest community.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) the soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific percent loss of leaf conductivity curves (PLC), and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf physiological traits collected during this study are described below, details of all other datasets employed are described in the Supplementary Information.

## The Study Site
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, including availability of an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit:1998395; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for diameter growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model increasingly used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher: ).

\
Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow (CLM4.5 Tech Note). ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow or wetlands and lakes), and can be described as

@@

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H2O) and $\Delta t$ is the time-step. z are the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for in Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 10, 000 ensembles from this parameter space with Latin Hypercube Sampling.

\
We ran ELM-FATES for these 10,000 ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

### Model evaluation
\
We calibrated the model against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration ($E_{v} + E_{g}$) from the eddy flux tower, (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$) and (iii) Volumetric Water Content (VWC) obtained with Time Domain Reflectometry (TDR) at three depths (10, 40 and 100 cm; relating to $W_{liq, z}$) at point location as well as opportunistic dry season VWC measured over those depths for the entire 50-ha plot (Figure \@ref(fig:hydro-calib). Among the 10,000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges.

```{r hydro-calib, dev='png', fig.show='hide', eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap ="\\label{lab-hydro-calib}Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensembles runs of ELM-FATES. This figure will be reformatted to compose individual figures into one."}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic forms of equations to species-level data as prescribed by [@Guyot:2011] and based on AIC criteria chose exponential form (Fig. S\@ref(fig:obs-plc-curves)), described as

@@

## Acknowledgements
## Data Availability
## Supplementary Information

>>>>

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom, kableExtra))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

@@

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~Stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{Leaf}$ vs. $\Psi_{Leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small[@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see [@Johnson:2018fad;@Kumagai:2011;@Ivanov:2012]

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~Stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity. A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~Stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~Stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~Stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting/water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs that shape how trees' make a living in their environment (ecological strategies) provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. A universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum conceptualises plants coordination for all resources and organs, both above- and below-ground [@Reich:2014]. For water resource observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high Kmax, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{TLP}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture gradient in the soil column is not clear. While deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high Kmax) that maximise the use of the water-pulse, while *deep*-rooted plants have slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of xx canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 35 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{x,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{x,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrined by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{Leaf}$ vs. $\Psi_{x,~Leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow trade-off axis. Specific hypotheses applicable to this forest included: (i) dynamics of soil water potential at depth does not cross critical plant thresholds even during multi-annual droughts because of high total annual precipitation to evapotranspiration ratio (not sure if this should be stated as a hypothesis or just an explanation of the results); (ii) trees with deeper ERD show 'fast' hydraulic traits: higher maximum stem hydraulic conductance (Kmax), but greater xylem vulnerability to embolism ($\Psi_{88,~Stem}$), higher $\Psi_{TLP}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~Stem}$); and finally, (iii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and mortality outcomes for a species-rich forest.

# Materials and Methods
\
In the following sections we describe the study site, the models and data used to estimate and evaluate (1) soil water potential dynamics by depth over the forests’ rooting zone, (2) species-specific leaf vulnerability curves, and (3) species-specific effective rooting depths. These are followed by analyses to address the proposed hypotheses. Note that only new data on leaf vulnerability curves collected during this study are described below, details of all other datasets used are described in the Supplementary Information.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterisation in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560 hectare island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD over 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50 ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50 ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2 ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].

\
Below we describe the equation of water balance in ELM-FATES that is most relevant for our study site, even though ELM-FATES is capable of incorporating numerous other hydrological processes relevant for wetlands and lakes and snow [@Bonan:2013]. ELM-FATES parameterizes interception, throughfall, canopy drip, infiltration, evaporation, surface runoff, subsurface drainage, redistribution within the soil column, and groundwater discharge and recharge so as to simulate changes in canopy water $\Delta W_{can}$, surface water $\Delta W_{sfc}$, soil water $\Delta W_{liq, z}$, and water in an unconfined aquifer $\Delta W_{a}$ (omitting processes relevant to snow, wetlands or lakes), and can be described as

@@

where, $q_{rain}$ is rainfall, $E_{v}$ is evapotranspiration (ET) of the forest, $E_{g}$ is ground evaporation, $q_{over}$ is surface runoff, $q_{h2osfc}$ is runoff from surface water storage, $q_{drai}$ is sub-surface drainage (all in mm H~2~O) and $\Delta t$ is the time-step. $z$ represents the 15 soil layers exponentially distributed in depth, here, (r soil.depths) (m).

### Model parameterization
\
To parameterise the model, we identified eleven parameters relevant for Eq. \@ref(eq:water-balance), determined their ranges based on literature for the study site or for the tropics ([Table 1](#param)), and sampled 5000 ensembles from this parameter space with Latin Hypercube Sampling (LHS). In the Conrad catchment soil hydraulic conductivity declined by depth based on existing data for the depths of 12.5 and 60 cm (@Kinner:2004; [Table 1](#param)]. We generated 5000 LHS samples within the 95% CI for the two depths and assumed a linear decline between each paired sample, while for depths < 12.5 cm and > 60 cm, we assumed the same value as that for depths 12.5 and 60 cm, respectively (Fig. S\@ref(fig:ksat-profiles)). For all runs, instead of ELM's default soil texture-based soil water retention curve scheme, we used parameters of a soil water retention curve that we derived from existing data on volumetric water content \theta vs. $\Psi_{soil}$ for depths 15, 40 and 100 cm [@Kupers:2019]. A Campbell (1974) empirical equation was fitted for all depths pooled together, as exploratory analyses indicated differences among depths were nominal:

\begin{equation}
\Psi_{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}

where $\Psi_e$ is the air-entry matric potential (200, mmH~2~O), \theta_{s} is the saturated volumetric water content (0.51, $cm^{3}cm^{-3}$), and $b$, an index for soil pore-size distribution (10, unitless), respectively.

We ran ELM-FATES for the 5000 parameter ensembles with hourly climate drivers from 1985-2018 initialised with the stand structure of the 50 ha plot in a single site mode ($0.1^\circ$ x $0.1^\circ$ centered at $9.1530^\circ$N, $280.2539^\circ$E). As our key interest here was in soil water availability rather than forest dynamics, we ran ELM-FATES in simplified modes: static stand structure and without plant hydrodynamics, and with a single Plant Functional Type (PFT) of evergreen trees. The latter was chosen as only 9.7% of BCI crown area is dry-season deciduous and addition of dry-deciduous PFT did not significantly alter results (not shown).

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib; daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm{-3} (n = 5, cylinder volume = 1250 cm{3}). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See [Table 1](#param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

```{r hydro-calib, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-hydro-calib} Model evaluation: Observations (red line) of Volumetric Water Content, Evapotranspiration and stream discharge against simulations (blue lines) from 100 best-fit parameter ensemble runs of ELM-FATES. This figure will be reformatted to compose individual figures into one.")}
p1 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/swc_Obs_vs_model_daily_horizontal_with_steph_mean.jpeg", scale = 1)
p2 <- cowplot::ggdraw() + cowplot::draw_image("figures/2019-10-14_5000/best-fits/QRUNOFF_Obs_vs_model_monthly.jpeg", scale = 1)
p3 <- cowplot::ggdraw() +
  cowplot::draw_image("figures/2019-10-14_5000/best-fits/ET_Obs_vs_model_monthly.jpeg", scale = 1)
bottom_row <- cowplot::plot_grid(p2, p3, labels = c('B', 'C'), label_size = 12, ncol = 2)
cowplot::plot_grid(p1, bottom_row, labels = c('A', ''), label_size = 12, ncol = 1, rel_widths = c(2, 1))
```

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{x,~Leaf}$ vs. $\Psi_{x,~Leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig. S\@ref(fig:obs-plc-curves)), described as

@@

## Acknowledgements
## Data Availability

#### Table 1: ELM-FATES parameters used to generate ensembles, with description, chosen ranges, references and rationale. No. 1-6 are parameters specific to FATES.  {#param}
